Categories of Optics

### 2018-06-11 paper.tex
More work

\section{Optics}

@@

@@

\begin{proposition}
If $F : \C \to \D$ is a functor that commutes with the monoidal actions on $\C$ and $\D$, then $\Optic(F) : \Optic_\M \to \Optic_\N$ restricts to a functor $\Lawful_\M \to \Lawful_\N$.
\end{proposition}
\begin{proof}
\todo{this should be in here}
\end{proof}

@@

@@

\section{Profunctor Optics}

@@

\begin{definition}
Suppose $\C$ is acted on by $(\M, \otimes, I)$ and let $P \in \Prof(\C, \C)$ be a profunctor. A \emph{Tambara module structure for $\M$} on $P$ is a family of maps:
\begin{align*}
\alpha_{A,B,M} : P(A,B) \to P(MA, MB)
\end{align*}
natural in $A$ and $B$, and dinatural in $M$. Further, we must have that $\alpha_{A,B,I} = \id_{P(A,B)}$, and that the following diagram commutes:
\[
\begin{tikzcd}
P(A,B) \ar[r, "\alpha_{A,B,M}"] \ar[dr, "\alpha_{A, B, N \otimes M}" below left] & P(MA, MB) \ar[d, "\alpha_{MA, MB, N}" right] \\
& P(NMA, NMB)
\end{tikzcd}
\]
for all $A, B \in \C$ and $N, M \in \M$.
\end{definition}

@@

This category is monoidal with respect to $\odot$ as given above with monoidal unit $I$. There is an evident forgetful functor $U : \Tamb(\M) \to \Prof(\C, \C)$ that is strong monoidal. This forgetful functor has both a left and right adjoint; important for us is the left adjoint:

@@

The data of such a map $\hat f : \Phi_\M P \hto T$ is a natural transformation between the underlying profunctors. For the factorisation property to hold, we must have that $\hat{f}\eta(p) = f(p)$ for any $p \in P(A,B)$, but then the action on the remainder of $\Phi_\M P(A, B)$ is fixed:
\begin{align*}
\hat{f}(\langle l, p, r) \rangle &= \hat{f}(\Pastro_\M P(l, r) \alpha_{A, B, M} \; \eta(p)) \\
&=T(l, r) \; \beta_{A, B, N} \; f(p)
\end{align*}
This establishes uniqueness. It remains to show that $\hat{f}$ so defined is actually a Tambara module morphism, but this is easy:
\begin{align*}
\hat{f}\alpha_{A,B,N}(\langle l, p, r \rangle)
&= \hat{f}(\langle Nl, p, Nr \rangle) \\
&= T(Nl, Nr) \; \beta_{A, B, NM} \; f(p) \\
&= T(Nl, Nr) \; \beta_{A, B, NM} \; f(p) \\
&= T(Nl, Nr) \beta_{MA,MB,N} \; \beta_{A, B, M} \; f(p) \\
&= \beta_{A,B,N} T(l, r) \; \beta_{A, B, M} \; f(p) \\
&= \beta_{A,B,N} \hat{f} (\langle l, p, r \rangle)
\end{align*}
\end{proof}

\begin{corollary}
$\Pastro_\M$ is oplax monoidal.
\end{corollary}
\begin{proof}
This follows from abstract nonsense as the left adjoint of a strong monoidal functor \cite{Kelly1974}.
\end{proof}

@@

\begin{proposition}[Profunctor Optics are Optics]
\label{prop-profunctor-optics-are-optics}
\begin{align*}
[\Tamb(\M), \Set]((U-)(A,A),(U-)(S,S)) &\cong \Optic_\M(S, A)
\end{align*}
\end{proposition}
\begin{proof}
We have the chain of isomorphisms:
\begin{align*}
&[\Tamb(\M), \Set]((U-)(A,A),(U-)(S,S)) \\
\cong \;&[\Tamb(\M), \Set](\Tamb(\M)(\Pastro_\M E_A, -), (U-)(S,S))  && \text{(by Lemma \ref{lemma-rep})}\\
\cong \;&(U\Pastro_\M E_A)(S,S)  && \text{(by Yoneda)} \\
= \;&\int^{M \in \M} \int^{C,D \in \C} \C(S, MC) \times E_A(C,D) \times \C(MD,S) \\
= \;&\int^{M \in \M} \int^{C,D \in \C} \C(S, MC) \times \C(C,A) \times \C(A,D) \times \C(MD,S) \\
\cong \;&\int^{M \in \M} \C(S, MA) \times \C(MA, S)  && \text{(by co-Yoneda twice)} \\
= \;&\Optic_\M(S, A)
\end{align*}
\end{proof}

The same proof works in the more general case of $\Optic_\M((S, S'), (A,A'))$, defining $E_{A, A'}$ to be $\C(-, A) \times \C(A', {=})$.

@@

@@

\subsection{Optics Into Stacks of Functors and Monad Transformers}

>>>>

\section{Optics}
\label{sec-optics}

@@

\begin{proposition}
If $F : \C \to \D$ is a functor that commutes with the monoidal actions on $\C$ and $\D$, then $\Optic(F) : \Optic_\M \to \Optic_\N$ restricts to a functor $\Lawful_\M \to \Lawful_\N$.
\end{proposition}
\begin{proof}
If $p = \langle l, r \rangle$ is lawful, 
\end{proof}

@@

@@

\subsection{Monad Transformer Optics}

\todo{todo}

There is also an analogous notion of optics of composites of functors.

@@

\section{Profunctor Optics}
The Haskell \texttt{lens} library implements optics using what is known as the \emph{profunctor encoding}. Lenses are defined as follows:

\begin{minted}{haskell}
class Profunctor p => Strong p where
  second' :: p a b -> p (c, a) (c, b)

type Lens s a = forall p. Strong p => p a a -> p s s
\end{minted}

The precise connection between this encoding and the previous one was investigated independently in \cite{ProfunctorOpticsPost}. In this section, we extend these ideas to the case of a monoidal action and include an analysis of lawful optics in this setting.

@@

\begin{definition}
Suppose a category $\C$ is acted on by $(\M, \otimes, I)$ and let $P \in \Prof(\C, \C)$ be a profunctor. A \emph{Tambara module structure for $\M$} on $P$ is a family of maps:
\begin{align*}
\alpha_{A,B,M} : P(A,B) \to P(MA, MB)
\end{align*}
natural in $A$ and $B$, and dinatural in $M$. Further, we must have that $\alpha_{A,B,I} = \id_{P(A,B)}$, and that the following diagram commutes:
\[
\begin{tikzcd}
P(A,B) \ar[r, "\alpha_{A,B,M}"] \ar[dr, "\alpha_{A, B, N \otimes M}" below left] & P(MA, MB) \ar[d, "\alpha_{MA, MB, N}" right] \\
& P(NMA, NMB)
\end{tikzcd}
\]
for all $A, B \in \C$ and $N, M \in \M$.
\end{definition}

@@

This category is monoidal with respect to $\odot$ as given above with monoidal unit $I$.

\begin{remark}
In the code, the \mintinline{haskell}{Strong} typeclass corresponds to supplying the structure map $\alpha$ for a profunctor \texttt{p}. Parametricity guarantees a form of (di)naturality for $M$, $A$ and $B$, but as with any definition in Haskell, the equations required of a Tambara module must be checked externally.
\end{remark}

There is an evident forgetful functor $U : \Tamb(\M) \to \Prof(\C, \C)$ that is strong monoidal. This forgetful functor has both a left and right adjoint; important for us is the left adjoint:

@@

The data of such a map $\hat f : \Phi_\M P \hto T$ is a natural transformation between the underlying profunctors. For the factorisation property to hold, we must have that $\hat{f}\eta(p) = f(p)$ for any $p \in P(A,B)$, but then the action on the remainder of $\Phi_\M P(A, B)$ is fixed:
\begin{align*}
\hat{f}(\langle l, p, r) \rangle &= \hat{f}(\Pastro_\M P(l, r) \alpha_{A, B, M} \; \eta(p)) \\
&=T(l, r) \; \beta_{A, B, N} \; f(p)
\end{align*}
This establishes uniqueness. It remains to show that $\hat{f}$ so defined is actually a Tambara module morphism, but this is easy:
\begin{align*}
\hat{f}\alpha_{A,B,N}(\langle l, p, r \rangle)
&= \hat{f}(\langle Nl, p, Nr \rangle) \\
&= T(Nl, Nr) \; \beta_{A, B, NM} \; f(p) \\
&= T(Nl, Nr) \beta_{MA,MB,N} \; \beta_{A, B, M} \; f(p) \\
&= \beta_{A,B,N} T(l, r) \; \beta_{A, B, M} \; f(p) \\
&= \beta_{A,B,N} \hat{f} (\langle l, p, r \rangle)
\end{align*}
\end{proof}

\begin{corollary}
$\Pastro_\M$ is oplax monoidal.
\end{corollary}
\begin{proof}
This follows from abstract nonsense as the left adjoint of a strong monoidal functor, see \cite{Kelly1974}.
\end{proof}

@@

\begin{proposition}[Profunctor Optics are Optics]
\label{prop-profunctor-optics-are-optics}
\begin{align*}
[\Tamb(\M), \Set]((U-)(A,A),(U-)(S,S)) &\cong \Optic_\M(S, A)
\end{align*}
\end{proposition}
\begin{proof}
We have the chain of isomorphisms:
\begin{align*}
&[\Tamb(\M), \Set]((U-)(A,A),(U-)(S,S)) \\
\cong \;&[\Tamb(\M), \Set](\Tamb(\M)(\Pastro_\M E_A, -), (U-)(S,S))  && \text{(by Lemma \ref{lemma-rep})}\\
\cong \;&(U\Pastro_\M E_A)(S,S)  && \text{(by Yoneda)} \\
= \;&\int^{M \in \M} \int^{C,D \in \C} \C(S, MC) \times E_A(C,D) \times \C(MD,S) \\
= \;&\int^{M \in \M} \int^{C,D \in \C} \C(S, MC) \times \C(C,A) \times \C(A,D) \times \C(MD,S) \\
\cong \;&\int^{M \in \M} \C(S, MA) \times \C(MA, S)  && \text{(by co-Yoneda twice)} \\
= \;&\Optic_\M(S, A)
\end{align*}
The same proof works in the more general case of $\Optic_\M((S, S'), (A,A'))$, defining $E_{A, A'}$ to be $\C(-, A) \times \C(A', {=})$.
\end{proof}

\begin{remark}
This proposition justifies the profunctor encoding
\begin{minted}{haskell}
type Lens s a = forall p. Strong p => p a a -> p s s
\end{minted}
using parametricity as a proxy for naturality.
\end{remark}

@@

Free copointed on $F$ is exactly $F \times \id_\C$
The functor $\mathsf{Copointed} \to [\C, \C]$ has a right adjoint given by $F \mapsto F \times \id_\C$. Copointed map $F \to G \times \id_\C$ is the same as an uncopointed map $F \to G$

For $l, r : \C(S, FA) \times \C(FA, S)$, the map $r$ is equal to
\[
FA \xrightarrow{[\id_{FA}, \mathsf{copoint}]} FA \times A \xrightarrow{\pi_1} FA \to S
\]
But that first map is the component at $A$ of a copointed natural transformation (the unit of the adjunction). So $l, r$ is equal to
\begin{align*}
S \to FA \xrightarrow{[\id_{FA}, \mathsf{copoint}]} FA \times A\\
FA \times A \xrightarrow{\pi_1} FA \to S
\end{align*}

@@
